\section{Introduction}

Rotary position embedding (RoPE; \citealp{su2024roformer}) is often discussed as a positional operator, but in implementation it is also a \emph{finite spectral allocation}: each attention head receives only $d_{\mathrm{head}}/2$ rotary frequency channels, and the default geometric (equal-ratio) schedule is one inherited convention for spending this budget. This convention propagates unchanged from LLaMA-3 \citep{grattafiori2024llama3} and Qwen2.5 \citep{qwen2024qwen25} to DeepSeek-V3 \citep{deepseekv3} and video DiTs \citep{videorope2025}, even as model scales, context lengths, and architectures vary. Once RoPE is viewed as a finite spectral budget, a basic question appears: before training begins, how should these channels be allocated?

The rotation is the operator; the frequency table is a design choice.

Most long-context work answers a different question. Base-frequency tuning changes the span of the spectrum; positional interpolation \citep{chen2024position}, NTK-aware scaling \citep{roziere2023codellama}, YaRN \citep{peng2024yarn}, and LongRoPE-style methods \citep{ding2024longrope,shang2025longrope2} rescale positions at or after inference. These \emph{range-scaling} methods stretch or reuse a frequency table whose allocation was fixed at initialization. Adjacent works observe that the allocation matters empirically---\citet{barbero2025round} find high- and low-frequency specialization after training, \citet{videorope2025} propose a modality-specific allocation for video, \citet{wang2024resonance} preserve critical frequencies during interpolation---but none gives a closed-form \emph{training-time} allocation for standard RoPE derived from an explicit variational surrogate.

We treat training-time frequency-channel allocation as a third RoPE design axis, complementary to operator and range. Parameterizing allocations by densities $\rho(\phi)$ over log-frequency, we introduce \textsc{EVQ-Cosh}: a closed-form one-parameter allocation family with zero learned parameters that changes only the RoPE inverse-frequency initialization. Geometric RoPE is recovered as the $\tau{\to}0$ limit of the pure-tether sub-family (not a generic optimum). A small-$\tau$ softmax-transport calculation motivates the deployable operating rule $\tau{=}d_{\mathrm{head}}/\sqrt{L}$, used as a robust default inside an empirically flat basin rather than a claim of global optimum.

\paragraph{Empirical scope.} We report three \emph{primary} stress tests (Primary~I--III in \S\ref{sec:experiments}, 3-seed where indicated): a matched-scale substrate test composing EVQ with YaRN on a $454$M transformer (3-seed); a PE-dominant $128{\to}8\mathrm{K}$ extrapolation test against DAPE \citep{zheng2024dape} (1--3 seed; the learned-PE row is 3-seed, EVQ/Geo/DAPE rows are seed 42); and a scarce-channel test on MLA \citep{deepseekv2, deepseekv3} with $16$ RoPE channels at $432$M (3-seed). We additionally report \emph{robustness} experiments on the same MHA setting, and \emph{supporting/exploratory} single- or two-seed evidence (video DiT, LoRA into LLaMA-3-8B, progressive training) that extends scope rather than anchoring primary claims; tier assignment appears in Table~\ref{tab:evidence-tier}.

These scope choices are part of the claim, not a retreat from it: EVQ does not replace inference-time scaling, but changes the training-time substrate on which methods such as YaRN act. We also do not claim universal downstream superiority; the target is PE-mechanism isolation at endpoints where positional allocation is causally stressed.

\paragraph{Contributions.}
\begin{itemize}
\item \textbf{Finite-spectral-budget view.} We identify RoPE training-time frequency-channel allocation as a third design axis, complementary to operator design and inference-time range scaling; geometric RoPE is an inherited allocation convention, not a consequence of the RoPE operator.
\item \textbf{Closed-form allocation family.} We introduce \textsc{EVQ-Cosh}: a closed-form one-parameter inverse-CDF allocation family with zero learned parameters, in which geometric RoPE is recovered as the $\tau{\to}0$ limit of the pure-tether sub-family. The method modifies only the inverse-frequency initialization.
\item \textbf{Deployable operating-point rule.} A small-$\tau$ softmax-transport argument gives the structural scaling $\tau^{*}\propto d_{\mathrm{head}}/\sqrt{L}$ with convention-dependent prefactor; $\tau{=}d_{\mathrm{head}}/\sqrt{L}$ is used as a robust default inside an empirically flat basin rather than a claimed global optimum.
\item \textbf{Three mechanism stress tests.} Matched-scale EVQ+YaRN tests substrate--range complementarity; PE-dominant extrapolation against DAPE tests whether allocation shape alone closes extrapolation gaps; compressed-RoPE MLA is the scarce-channel regime where allocation quality should matter most.
\end{itemize}
