\section{Related Work}

Positional encodings can be separated along three axes: the \emph{operator} that injects positional information, the \emph{inference-time} mechanism that extends its usable range, and \emph{training-time frequency allocation} / frequency-usage analysis of what channels actually do after training. We position EVQ-Cosh in each axis below.

\paragraph{Axis A: Positional-encoding operators.}
Transformers rely on positional structure to recover order \citep{vaswani2017attention}; the design space spans absolute encodings, relative biases \citep{shaw2018self, raffel2020exploring}, additive linear biases (ALiBi; \citealp{press2022alibi}), kernelized biases (KERPLE; \citealp{chi2022kerple}), rotary position embedding (RoPE; \citealp{su2024roformer}), magnitude-decay rotary (XPOS; \citealp{sun2022xpos}), learned positional functions (FIRE, \citealp{li2024fire}; DAPE, \citealp{zheng2024dape}), and operator variants such as HoPE/hyperbolic rotary encoding \citep{dai2025hyperbolicrope} and PaTH \citep{yang2025pathattention}. EVQ-Cosh is not a new operator: it preserves the RoPE rotation and uses cosh only as the closed-form solution of a variational ODE on log-frequency density---modifying how the fixed number of frequency channels is distributed along the spectrum. In the PE-dominant diagnostic (\S\ref{sec:exp-pe}) this zero-learned-parameter allocation change attains lower seed-42 extrapolation PPL than the DAPE-style learned positional-operator baseline, isolating the frequency-structure effect from operator capacity in that tested protocol.

\paragraph{Axis B: Inference-time and post-training context extension.}
NTK-aware scaling \citep{roziere2023codellama}, positional interpolation \citep{chen2024position}, YaRN \citep{peng2024yarn}, and base-limit analyses \citep{xu2024base} rescale or bound RoPE positions; in particular, base-limit analyses study how the RoPE base bounds attainable context length, whereas EVQ holds the base fixed and reallocates finite channels within that span. SelfExtend \citep{jin2024selfextend} uses grouped attention without fine-tuning; LongRoPE \citep{ding2024longrope} and PoSE \citep{zhu2024pose} use progressive or skip-wise strategies; LongRoPE2 \citep{shang2025longrope2} searches per-channel scaling factors by evolution on a frozen checkpoint and combines this with mixed-context-window adaptation; CLEX \citep{chen2024clex} trains a continuous extrapolation module. These methods start from a fixed RoPE allocation and focus on range extension, post-hoc rescaling, or adaptation after that allocation is fixed. EVQ-Cosh is orthogonal: it changes the \emph{training-time} frequency substrate itself, and in our experiments (\S\ref{sec:exp-yarn}) composes with YaRN under matched YaRN scale. \emph{Relation to LongRoPE2.} LongRoPE2 is the closest post-hoc per-channel rescaling method to EVQ, but the two target different stages: LongRoPE2 searches scaling factors for a pretrained checkpoint, whereas EVQ specifies the training-time frequency allocation before optimization begins. The methods are therefore conceptually composable; we leave empirical EVQ+LongRoPE2 composition to future work.

\paragraph{Axis C: Frequency-usage and allocation analysis.}
\citet{barbero2025round} show empirically that high-frequency channels specialize in positional patterns while low-frequency channels carry semantic content, an asymmetry that geometric allocation does not model. \citet{wang2024resonance} (Resonance RoPE) identify critical frequencies that should not be perturbed during interpolation, implicitly acknowledging channel inequality but framed around interpolation rather than allocation. \citet{chen2025hope} replace specific RoPE components with position-independent ones while retaining high-frequency rotary signals for extrapolation; EVQ instead preserves the full RoPE rotation and reallocates the same finite channel budget by a closed-form training-time density. \citet{videorope2025} propose sequence-length-dependent allocation tailored for video generation; \citet{zhao2025riflex} correct frequencies at inference time for video diffusion; \citet{qiu2024freenoise} address temporal coherence via noise rescheduling (adjacent, not a positional-encoding method). Concurrent work covers adjacent facets: MHRoPE \citep{qwen2026mhrope} and VLM-HoPE \citep{li2025hope} design or reassign axis-/modality-specific allocations for multimodal RoPE in vision--language models; FoPE \citep{hua2025fope} models each channel as a Fourier series; CARoPE \citep{veisi2025carope} learns dynamic per-head frequencies; Clipped RoPE \citep{li2026copeclipped} soft-clips low frequencies post-hoc. To our knowledge, none derives a closed-form training-time spectral density for standard RoPE from a variational principle.

\paragraph{Evaluation for long-context behavior.}
Long-context evaluation spans synthetic retrieval diagnostics and broader task suites \citep{bai2024longbench, hsieh2024ruler}. Prior work shows that long contexts can fail even when relevant information is nominally present \citep{liu2024lost}, and that positional encoding modifications can mitigate such failures \citep{zhang2024found}. We use both PE-dominant settings (isolating frequency-allocation effects) and retrieval-heavy evaluation (testing systems-level claims), treating incomplete larger-scale evidence conservatively.

\paragraph{Positioning.} EVQ-Cosh targets the unfilled cell of a training-time, closed-form, zero-learned-parameter frequency allocation for standard RoPE before optimization begins. It does not compete with range-extension methods by offering another rescaling rule, nor with learned PE methods by adding operator capacity; it changes the pretraining substrate---the frequency table that subsequent RoPE-based range-scaling and operator-adaptation methods inherit. Per-channel learned/searched methods \citep{shang2025longrope2, veisi2025carope, hua2025fope} target distinct stages or parameterizations, making the families stage-distinct and potentially composable; empirical head-to-head benchmarking is left to future work. A compact method comparison is provided in the appendix; EVQ-Cosh occupies the training-time, closed-form, zero-parameter allocation cell.
