\section{Predictions and Mechanism}

The theory predicts an asymmetric tradeoff. EVQ does not ``push frequencies to the low end''; instead it redistributes channel density, expanding low-frequency spacing while compressing high-frequency spacing. The anticipated gain is therefore concentrated in the regime where long-range discrimination is bottlenecked by low-frequency collisions.

This leads to a waterbed interpretation. Short-range costs should remain bounded because high-frequency channels are redundant for nearby positions, while long-range gains can be large because the low-frequency bottleneck is the dominant failure mode. The same picture predicts a base dependence: if frequencies with wavelength larger than \(L_{\mathrm{train}}\) define a collision block of relative size \(c \approx \ln L / \ln b\), then EVQ should help most when this block is large (high base, short \(L\)) and help much less when it is already tiny.

The same mechanism predicts complementarity with inference-time scaling. YaRN rescales positions at inference time, compressing usable frequency spacing. If the training-time allocation is already geometric, this compression acts on an already crowded low-frequency regime. EVQ expands that spacing before YaRN is applied, which is why the strongest systems result is EVQ combined with YaRN rather than EVQ in isolation.

Finally, the theory predicts that EVQ should \emph{not} create gains when the collision-block analysis predicts little room for improvement. The low-base dead-zone experiments confirm this: a negative result in that regime is evidence that the mechanism is specific rather than overfit.
