\section{Limitations and Scope}

\paragraph{Theory-to-practice gap.} The variational derivation (ODE, closed-form family, waterbed) is exact conditional on the broadband surrogate $K_{\mathrm{app}}$, validated across 12 configurations (Appendix~\ref{sec:surrogate-validation}). The practical $\tau^*$ scaling law is semi-analytic: $d_{\mathrm{head}}$ dependence and the $L^{-1/2}$ exponent are derived, but one transport constant $\lambda$ is calibrated empirically. We do not claim a fully first-principles closed formula. The observed $<1\%$ PPL gap across 27 configurations indicates that the remaining theoretical headroom is small, but closing $\lambda$ analytically is an open question.

\paragraph{Plug-and-play boundary.} Zero-shot frequency replacement on Qwen2.5-7B-Instruct (21-task LongBench, 5000-sample bootstrap, FDR correction) yields $-0.35\%$ overall (not significant): retrieval gains are offset by multi-hop QA losses. This is expected---changing allocation without adaptation disrupts learned attention patterns. The LoRA result (Table~\ref{tab:lora-8b}) demonstrates that even 300~steps of adaptation recovers $8$--$19\times$ extrapolation gains, establishing a practical deployment path. EVQ is best understood as a training-time initialization, not a zero-shot plug-in for pretrained models.

\paragraph{Scale and evaluation.} From-scratch experiments span five scales (50M--750M text) and video DiT (129M--382M); LoRA injection extends validation to 7B. Full from-scratch training at ${\geq}1\mathrm{B}$ is deferred to future work, though consistent improvement across tested scales suggests the mechanism is scale-independent. The progressive result, the video DiT 382M scale-up, and the MLA 1B-token YaRN fine-tuning composition are each single-seed; all are labeled as supporting evidence. Multi-seed replication of the progressive and MLA composition results is in progress.

\paragraph{Training-time method, not zero-shot plug-in.} EVQ modifies the frequency initialization before training. It is not a zero-shot replacement for pretrained models' frequency schedules. The Qwen2.5-7B zero-shot result ($-0.35\%$, not significant) confirms this boundary; even 300 steps of LoRA adaptation suffices to recover large extrapolation gains (Table~\ref{tab:lora-8b}).

\paragraph{LoRA regime.} The variational analysis predicts a phase transition at $r_c = d_{\mathrm{head}}/2$ (Appendix~\ref{sec:lora-regime}); empirically, LoRA at $r{=}64{=}d_{\mathrm{head}}/2$ recovers $88$--$95\%$ extrapolation PPL reduction (Table~\ref{tab:lora-8b}), consistent with the prediction. RULER scores do not improve, indicating LoRA rank suffices for positional adaptation but not instruction-following at extrapolated lengths; longer adaptation or full-rank fine-tuning may close this gap.
