\section{Conclusion}

We cast RoPE frequency allocation as a variational inverse problem and derive EVQ-cosh, a closed-form family in which geometric RoPE is the degenerate \(\tau{=}0\) limit. EVQ improves extreme extrapolation and, more importantly, unlocks inference-time scaling: EVQ+YaRN substantially outperforms Geo+YaRN, extending functional context to \(48\mathrm{K}\) (\(24\times\) training length). On Multi-head Latent Attention, where only 16 frequency channels carry positional information, EVQ alone outperforms Geo+YaRN at \(2\text{--}3\times\) extrapolation. Preliminary cross-modal transfer to video DiT provides supporting evidence that the same frequency-allocation mechanism extends beyond text.

The practical lesson is that long-context systems depend not only on inference-time scaling rules, but crucially on a better training-time frequency substrate. EVQ is a one-line initialization change---it modifies the inverse-frequency schedule, adds zero parameters, and composes with existing inference-time methods---yet on MLA it outperforms Geo+YaRN---the strongest tested inference-time baseline---at \(2\text{--}3\times\) extrapolation. The progressive amplification finding further suggests that teams already using staged context extension receive EVQ's benefit for free, with growing returns at each stage.

More broadly, the frequency-allocation axis has remained essentially unchanged since RoFormer~\citep{su2024roformer}, while inference-time scaling has received intense optimization. Our results suggest these two axes are comparably important but far from equally explored. The variational framework provides a principled starting point: the waterbed inequality bounds what any allocation can achieve, and the \(\tau^*\) scaling law eliminates hyperparameter search. As architectures compress RoPE into fewer channels---16 in MLA, potentially fewer in future designs---each channel's placement becomes proportionally more critical, making principled allocation a prerequisite rather than a refinement.
