\section{Proof Details}
\label{sec:proofs}

\subsection{From the broadband surrogate to the ODE}

Starting from the surrogate kernel
\begin{equation}
K_{\mathrm{app}}(\phi,\psi)=\alpha\delta(\phi-\psi)+\beta\min(\phi,\psi),
\end{equation}
the functional is
\begin{equation}
\mathcal{J}[\rho]
=
\frac{\alpha}{2}\int_0^1 \rho(\phi)^2\,d\phi
+
\frac{\beta}{2}\int_0^1\int_0^1 \rho(\phi)\rho(\psi)\min(\phi,\psi)\,d\phi\,d\psi
- \mu\int_0^1 \rho(\phi)b^{-2\phi}\,d\phi.
\end{equation}
Its first variation is
\begin{equation}
\delta\mathcal{J}
=
\int_0^1
\left[
\alpha\rho(\phi)
+
\beta\int_0^1 \rho(\psi)\min(\phi,\psi)\,d\psi
- \mu b^{-2\phi}
\right]\delta\rho(\phi)\,d\phi.
\end{equation}
The Euler-Lagrange stationarity condition is therefore
\begin{equation}
\alpha\rho(\phi)
+
\beta\int_0^1 \rho(\psi)\min(\phi,\psi)\,d\psi
=
\mu b^{-2\phi}.
\end{equation}
Define
\begin{equation}
g(\phi)=\int_0^1 \rho(\psi)\min(\phi,\psi)\,d\psi
=
\int_0^\phi \rho(\psi)\psi\,d\psi
+
\phi\int_\phi^1 \rho(\psi)\,d\psi.
\end{equation}
Direct differentiation gives
\begin{align}
g'(\phi)
&=
\int_\phi^1 \rho(\psi)\,d\psi,\\
g''(\phi)
&=
-\rho(\phi).
\end{align}
Hence the stationarity equation can be rewritten as
\begin{equation}
\alpha \rho(\phi)+\beta g(\phi)=\mu b^{-2\phi}.
\end{equation}
Differentiating twice with respect to \(\phi\) and substituting \(g''(\phi)=-\rho(\phi)\) yields
\begin{equation}
\rho''(\phi)-\frac{\beta}{\alpha}\rho(\phi)=\gamma b^{-2\phi},
\qquad
\tau=\sqrt{\beta/\alpha}.
\end{equation}
The same calculation shows that \(g(0)=0\) and \(g'(1)=0\), so \(\min(\phi,\psi)\) acts as the Green kernel of \(-d^2/d\phi^2\) under the mixed boundary conditions induced by the surrogate formulation. Conditional on the surrogate, this step is exact.

\subsection{Closed-form solution and geometric limit}

The non-resonant ODE has the closed-form solution
\begin{equation}
\rho^*(\phi)=C_1\cosh(\tau\phi)+C_2\sinh(\tau\phi)+P\,b^{-2\phi}.
\end{equation}
The first two terms span the homogeneous solution of \(\rho''-\tau^2\rho=0\), and the final term is a particular solution induced by the Fisher-weighted source \(b^{-2\phi}\). This decomposition is the source of the main paper's ``tether + pulse'' interpretation.

To translate the density into RoPE frequencies, write the cumulative map
\begin{equation}
F(\phi)=\int_0^\phi \rho^*(s)\,ds,
\end{equation}
normalize it to \([0,1]\), and invert at the discrete quantiles \(u_k=k/K\). In the pure-tether branch this inversion has the closed form
\begin{equation}
\phi_k(\tau)
=
1-\frac{1}{\tau}\operatorname{arcsinh}\!\left((1-u_k)\sinh\tau\right).
\end{equation}
To recover geometric RoPE, expand \(\sinh\tau\approx\tau\) as \(\tau\to 0\). Then
\begin{equation}
\phi_k(\tau)\to 1-\frac{1}{\tau}\operatorname{arcsinh}((1-u_k)\tau)\to u_k,
\end{equation}
which is exactly the uniform quantile map.

This also shows why the high-frequency end remains effectively anchored: for \(u_0=0\), one has \(\phi_0(\tau)=0\) exactly, so the leading inverse frequency remains \(b^0=1\), matching geometric RoPE.

\subsection{Why the pure-tether branch}

The full stationary density contains three terms: a hyperbolic tether (\(\cosh, \sinh\)) and a particular solution \(Pb^{-2\phi}\). The practical EVQ method uses only the pure-tether branch. This is a deliberate design choice, not an oversight, and is justified on three grounds.

\textbf{(1) Separation of concerns.} The homogeneous solution captures the global allocation structure --- how frequency mass should be distributed across the spectrum to minimize broadband collision. The particular solution \(Pb^{-2\phi}\) captures a local Fisher-utility correction that preserves the geometric profile's per-channel information content. In the regime where collision reduction is the binding constraint (i.e., when many frequency channels are wasted by the geometric allocation), the global allocation term dominates the quality improvement, and the local correction is second-order.

\textbf{(2) Closed-form invertibility.} The pure-tether density \(\rho(\phi)\propto\cosh(\tau(1-\phi))\) admits an analytically invertible CDF, yielding the closed-form warp \(\phi_k(\tau)\). Including the \(b^{-2\phi}\) term destroys this invertibility and requires numerical root-finding, sacrificing the method's main practical advantage: a deterministic, parameter-free initialization formula.

\textbf{(3) The forcing term is amplitude-suppressed.} The particular solution coefficient $P = \gamma/(4\ln^2 b - \tau^2)$ is strongly suppressed at typical base values: $P \propto 1/687$ at $b{=}500\mathrm{K}$ and $1/337$ at $b{=}10\mathrm{K}$. We verify numerically by computing collision scores under the exact kernel for the full density $\rho(\phi) = \cosh(\tau(1{-}\phi)) + \lambda b^{-2\phi}$ across $\lambda \in [0, 5]$. At the theoretically motivated $\lambda \approx P \approx 0.001$--$0.003$, the full solution improves over pure tether by $<0.5\%$. Even at artificially large $\lambda{=}1$, the improvement is only $2$--$6\%$ across configurations (vs.\ the $28$--$92\%$ improvement over geometric). The pure tether already captures the vast majority of the collision reduction.

\textbf{(4) Empirical validation of the design choice.} The pure-tether branch subsumes geometric RoPE as its \(\tau\to 0\) limit, and the 99-run Phase~16 sweep (Fig.~6) confirms that the optimal \(\tau\) in this family matches the scaling-law prediction across multiple model scales. If the omitted forcing term were essential, one would expect the pure-tether family to systematically underperform --- but in fact it consistently improves over geometric at the predicted \(\tau^*\). The forcing term may further improve performance, but this is an avenue for future work rather than a deficiency of the current method.

\subsection{Surrogate quality: functional validation}
\label{sec:surrogate-validation}

The broadband surrogate $K_{\mathrm{app}} = \alpha\delta + \beta\min$ is the theory's sole approximation. We validate it not by pointwise comparison with the exact kernel $K(\phi_1,\phi_2) = \int D(\Delta)\cos(\omega(\phi_1)\Delta)\cos(\omega(\phi_2)\Delta)\,d\Delta$ --- which is oscillatory (sinc-like) and cannot be well-approximated pointwise by a smooth two-parameter model --- but by the \emph{functional} test: does the EVQ allocation derived from $K_{\mathrm{app}}$ also reduce the collision objective under the true kernel $K$?

Table~\ref{tab:surrogate-validation} answers affirmatively. For a uniform distance prior $D(\Delta) = 1/L$ on $[0, L]$, we compute the exact kernel $K$ at the discrete allocation points for both geometric and EVQ, then evaluate the normalized collision score $C = \sum_{i<j} K_{ij}^2/(K_{ii}K_{jj})$ and effective spectral rank (exponential of the Shannon entropy of the eigenvalue distribution). We sweep across six training lengths ($L \in \{128, \ldots, 4096\}$), three base values ($b \in \{10\mathrm{K}, 100\mathrm{K}, 500\mathrm{K}\}$), and two channel counts ($K \in \{16, 32\}$), covering both text and video regimes.

\begin{table}[ht]
\caption{Functional surrogate validation across 12 configurations. EVQ (derived from $K_{\mathrm{app}}$) reduces collision and increases effective rank under the \emph{exact} kernel $K$ in every tested regime. Collision reduction monotonically increases with extrapolation stress (shorter $L$, larger $b$).}
\label{tab:surrogate-validation}
\centering
\small
\begin{tabular}{@{}l r r r r r@{}}
\toprule
& \multicolumn{2}{c}{Collision $C$} & & \multicolumn{2}{c}{Eff.\ rank} \\
\cmidrule(lr){2-3} \cmidrule(lr){5-6}
Configuration & GEO & EVQ ($\Delta_C$) & & GEO & EVQ \\
\midrule
\multicolumn{6}{@{}l}{\emph{Text: $K{=}32$, $b{=}500\mathrm{K}$, sweep $L$}} \\
\quad $L{=}128$ & 226.3 & 17.7 ($-92\%$) & & 2.5 & 17.0 \\
\quad $L{=}256$ & 192.6 & 27.6 ($-86\%$) & & 3.0 & 13.9 \\
\quad $L{=}512$ & 163.2 & 43.3 ($-73\%$) & & 3.6 & 10.8 \\
\quad $L{=}1024$ & 133.9 & 59.0 ($-56\%$) & & 4.4 & 8.8 \\
\quad $L{=}2048$ & 109.0 & 67.5 ($-38\%$) & & 5.4 & 8.0 \\
\quad $L{=}4096$ & 87.0 & 66.3 ($-24\%$) & & 6.6 & 8.1 \\
\midrule
\multicolumn{6}{@{}l}{\emph{Text: $K{=}32$, $L{=}2048$, sweep $b$}} \\
\quad $b{=}10\mathrm{K}$ & 34.8 & 19.2 ($-45\%$) & & 13.0 & 17.3 \\
\quad $b{=}100\mathrm{K}$ & 79.4 & 47.0 ($-41\%$) & & 7.1 & 10.4 \\
\midrule
\multicolumn{6}{@{}l}{\emph{Video: $K{=}16$, $L{=}32$, sweep $b$}} \\
\quad $b{=}100$ & 22.6 & 14.2 ($-37\%$) & & 4.8 & 6.4 \\
\quad $b{=}1\mathrm{K}$ & 42.3 & 28.0 ($-34\%$) & & 2.8 & 3.9 \\
\quad $b{=}10\mathrm{K}$ & 55.4 & 39.8 ($-28\%$) & & 2.2 & 2.9 \\
\quad $b{=}50\mathrm{K}$ & 63.3 & 47.2 ($-26\%$) & & 2.0 & 2.5 \\
\bottomrule
\end{tabular}
\end{table}

EVQ reduces collision under the exact kernel in all 12 configurations ($-24\%$ to $-92\%$), with effective rank improvements of $+24\%$ to $+570\%$. Two patterns emerge. First, the collision reduction monotonically increases as the geometric allocation becomes more degenerate: shorter $L$ (more extrapolation stress at fixed $\tau^*$) and larger $b$ (more dead channels) both amplify the advantage, consistent with the dead-channel mechanism. Second, even in the most favorable regime for geometric RoPE ($L{=}4096$, where baseline effective rank is already $6.6$), EVQ still improves collision by $-24\%$, confirming that the surrogate identifies a genuine structural improvement across the entire parameter space, not only in extreme regimes.

\subsection{Waterbed assumptions}

The waterbed proposition in the body is conditional. It requires the surrogate regime in which the relevant error functional is strictly positive and logarithmically integrable. Under those assumptions, Jensen's inequality yields a lower bound of the form
\begin{equation}
\int \ln E(\phi)\,d\phi \ge \ln b - \ln c,
\end{equation}
with equality at the geometric allocation. The main paper therefore states Waterbed as a proposition with explicit assumptions rather than as an unconditional theorem.

The logic is the standard convexity argument. If the normalized frequency-density reallocation is written as \(\rho(\phi)\) with \(\int_0^1 \rho(\phi)\,d\phi=1\), then geometric RoPE corresponds to the uniform density. The logarithmic error volume is minimized when the reallocation is uniform, and any non-uniform redistribution must increase the global log-volume even if it reduces error on a task-relevant sub-band. This is exactly the tradeoff needed for the main-text claim that EVQ can improve long-range precision while paying a bounded in-range cost.

\subsection{Why the scaling law remains a conjecture}

The current argument for
\begin{equation}
\tau^*(L)=\frac{d_{\mathrm{head}}}{\sqrt{L}}
\end{equation}
combines asymptotic scaling intuition with experimental confirmation across multiple \(L\) values. What is still missing is a theorem that turns the scale analysis into an explicit bound with constants, so the body keeps this statement at the level of an empirical law / conjecture.

In particular, the current derivation argues that the surrogate coefficients satisfy an effective scaling of the form
\begin{equation}
\alpha^*(L,b)\propto L\ln b,
\qquad
\beta^*(L,b)=O(1).
\end{equation}
The rationale is that the diagonal penalty \(\alpha\) reflects per-channel position resolution, which grows with both the sequence length \(L\) (more position tokens to distinguish) and the log-base \(\ln b\) (wider frequency spread). The off-diagonal coupling \(\beta\) captures the integrated cross-channel interference structure, which is set by the surrogate's Green kernel and does not scale with \(L\). This gives
\begin{equation}
\tau^*(L,b)=\sqrt{\frac{\beta^*(L,b)}{\alpha^*(L,b)}}\propto \frac{1}{\sqrt{L\ln b}}.
\end{equation}
Fixing \(b\) then yields the single-variable form used in the paper. The empirical evidence is strong enough to justify using this law as the practical recipe, but not yet strong enough to promote it from conjecture to theorem. That distinction is maintained explicitly in the body.
