\section{Experimental Details}

\subsection{Primary body experiments}

The body centers three empirical anchors:
\begin{enumerate}
\item the EVQ\(\times\)YaRN systems result on passkey-mix data,
\item DAPE-style PE-dominant extrapolation at \(128\) tokens,
\item the \(L=256\) Phase 11 confirmation with raw and YaRN curves.
\end{enumerate}
These anchors were selected because they are either multi-seed or directly tied to the central theoretical prediction.

\subsection{Reproducibility snapshot}

\begin{table}[ht]
\caption{Compact reproducibility summary for the primary experiments discussed in the main text.}
\label{tab:appendix-repro}
\centering
\small
\begin{tabular}{@{}lllll@{}}
\toprule
Anchor & Train length & Scale & Seeds & Main datasets \\
\midrule
EVQ\(\times\)YaRN & 2048 & 350M & 3 + 3 & FineWeb-Edu + passkey mix \\
DAPE-style & 128 & 125M & 1--3 & FineWeb-Edu / TinyStories \\
Phase 11 & 256 & 454M & 3 & FineWeb-Edu \\
\bottomrule
\end{tabular}
\end{table}

Across all three anchors, the architectural intervention is deliberately minimal: model architecture, optimizer family, and training loop are kept fixed, and only the RoPE inverse-frequency initialization is changed. The principal base in the body is \(b=500\mathrm{K}\), which is also the regime where the collision-block analysis predicts the largest headroom for non-geometric allocations.

\subsection{Larger-scale supporting evidence}
\label{sec:750m-supporting}

The \(750\mathrm{M}\) `continue@4K' experiment is reported as supporting evidence only. The reason is not that the result is weak; on the contrary, it shows \(16\mathrm{K}\) PPL improving from \(45.136\) to \(24.407\) and \(8\mathrm{K}\) AR exact improving from \(0\%\) to \(77.5\%\). The reason is evidence hygiene: the run is single-seed, and the downstream LongBench comparison is incomplete because the remote dataset download failed during evaluation.

This distinction is important for interpretation. The larger-scale run is best viewed as evidence that the same EVQ pattern persists after continued pretraining at a longer training length, not as the main fairness anchor of the paper. The main fairness anchors remain the multi-seed \(350\mathrm{M}\) and \(454\mathrm{M}\) comparisons used in the main text.

\input{tables/table6_750m_continue_supporting}

\subsection{Evaluation notes}

The paper uses three increasingly strict long-context evaluation views.
\begin{enumerate}
\item \textbf{PPL}: global language-modeling quality under extrapolation.
\item \textbf{Retrieval}: whether the relevant item is still localized at long range.
\item \textbf{AR exact / multi-needle}: whether the localized information remains precisely recoverable.
\end{enumerate}

This split is necessary because stronger models can saturate coarse retrieval while still differing sharply in exact long-range recovery. The \(750\mathrm{M}\) supporting run is the clearest example: retrieval is already saturated, but AR exact separates geometric RoPE from EVQ very strongly.

\subsection{Modality-dependent $\tau$ correction for video DiT}
\label{sec:tau-correction}

The scaling law $\tau^*=d_{\mathrm{head}}/\sqrt{L}$ matches text autoregressive models with causal attention. For bidirectional DiT, the empirically optimal $\tau$ is approximately $0.53\times$ the text prediction ($\tau{=}1.5$ vs.\ predicted $2.83$ for $K_t{=}16, T{=}32$). This ratio reflects at least two compounding factors. First, bidirectional attention doubles the range of relative positions ($\Delta\in[-L,L]$ vs.\ $[0,L]$), predicting a correction of $1/\sqrt{2}\approx 0.71$. Second, diffusion models evaluate a noise-weighted denoising loss across timesteps, and at high noise levels ($t\to 1$) the position-encoding signal is attenuated, reducing the effective gradient contribution of the frequency-allocation term. The product (${\approx}0.71 \times 0.75 \approx 0.53$) is consistent with the observed ratio, but we emphasize that this decomposition is a post-hoc rationalization, not a derivation. A principled modality-aware $\tau^*$ formula remains future work.

\subsection{Cross-modal evidence: video temporal extrapolation}

Video temporal extrapolation tests whether the same frequency-allocation principle transfers beyond text RoPE. We apply EVQ-Cosh to the temporal channels of a 3D RoPE VideoGPT (268.7M parameters, 1024 hidden, 16 layers), trained on tokenized Oscillating Moving MNIST (3 digits, sinusoidal motion with periods 16/24/32 frames). Models train on 32-frame sequences and extrapolate to 128 frames (4$\times$) with temporal YaRN scaling.

\paragraph{Setup.} We compare \texttt{geo\_k16} (16 temporal channel pairs, geometric allocation, $\tau{=}0$) against \texttt{evq\_k16} (same budget, EVQ-Cosh with $\tau{=}K_t/\sqrt{T_\mathrm{train}}$). Both allocations share identical spatial channels. Models are evaluated on PPL, teacher-forced top-1 accuracy, and generation quality via Fr\'echet Video Distance (FVD). For FVD, we report both I3D features and VideoMAE-v2 features; the latter are approximately $3\times$ more sensitive to temporal distortion than I3D~\citep{ge2024contentbiasfvd}.

\paragraph{Results.} Table~\ref{tab:video-temporal} summarizes the findings. EVQ wins every metric, but the margins attenuate sharply from distributional quality (PPL) through generation:

\begin{table}[ht]
\caption{Signal attenuation from distributional quality to generation metrics on Oscillating Moving MNIST (128 frames, 4$\times$ extrapolation). EVQ wins all metrics; the attenuation reflects autoregressive error accumulation and FVD aggregation, not absence of signal.}
\label{tab:video-temporal}
\centering
\small
\begin{tabular}{@{}llrrr@{}}
\toprule
Stage & Metric & geo\_k16 & evq\_k16 & $\Delta$ \\
\midrule
Distribution & PPL (YaRN) & 4.787 & \textbf{3.481} & $-27.3\%$ \\
Distribution & PPL (raw) & 4.832 & \textbf{4.440} & $-8.1\%$ \\
Prediction & Teacher-forced top-1 & 0.708 & \textbf{0.730} & $+3.1\%$ \\
Generation & VideoMAE FVD $\downarrow$ & 70.25 & \textbf{69.18} & $-1.5\%$ \\
Generation & I3D FVD $\downarrow$ & 239.9 & \textbf{238.7} & $-0.5\%$ \\
\bottomrule
\end{tabular}
\end{table}

The $3.2\times$ sensitivity ratio between VideoMAE and I3D FVD is consistent with the finding that I3D features are dominated by per-frame spatial content and nearly blind to temporal structure~\citep{ge2024contentbiasfvd}. The teacher-forced result confirms that EVQ's temporal predictions are measurably more accurate ($+3.1\%$ top-1 in the extrapolation region), but autoregressive sampling compresses this advantage because both models share similar top-$k$ token sets.

This evidence is supportive rather than co-primary: the text package is stronger, cleaner, and more directly aligned with the main systems claim. Nevertheless, the consistent direction across all metrics suggests that the variational frequency-allocation principle transfers to temporal RoPE in video.

\paragraph{Cross-architecture validation: DiT with bidirectional attention.}
To test whether EVQ transfers beyond autoregressive decoders, we evaluate on a 129.6M video diffusion transformer (DiT) with bidirectional attention and 3D RoPE. We compare EVQ-Cosh ($\tau{=}1.5$) against geometric allocation ($\tau{=}0$) on Oscillating Moving MNIST, training on 32 frames and extrapolating to 128 frames ($4\times$) with temporal YaRN. Crucially, both allocations are trained \emph{within the same run} (head-to-head), sharing identical weights, optimizer state, and CUDA non-determinism---eliminating all confounders except the frequency allocation itself.

Table~\ref{tab:dit-h2h} reports results across two independent seeds. EVQ wins every metric in both seeds. Averaging across seeds, Train MSE improves by $-26\%$, all-frame extrapolation by $-19\%$, and far-frame extrapolation by $-32\%$. The consistency across both seeds---EVQ wins all three metrics in each---suggests that the advantage is not a stochastic artifact, though we note the sample size (two seeds) is too small for formal statistical testing.

\begin{table}[ht]
\caption{DiT head-to-head evaluation (129.6M, bidirectional attention, 3D RoPE). Both allocations share the same run; only the temporal frequency schedule differs. EVQ wins all metrics across both seeds.}
\label{tab:dit-h2h}
\centering
\small
\begin{tabular}{@{}l l r r r@{}}
\toprule
Seed & Metric (YaRN) & Geo ($\tau{=}0$) & EVQ ($\tau{=}1.5$) & $\Delta$ \\
\midrule
42 & Train MSE & 0.01350 & \textbf{0.01069} & $-21\%$ \\
42 & All extrap & 0.00670 & \textbf{0.00573} & $-14\%$ \\
42 & Far extrap & 0.00774 & \textbf{0.00502} & $-35\%$ \\
\midrule
137 & Train MSE & 0.01113 & \textbf{0.00761} & $-32\%$ \\
137 & All extrap & 0.00713 & \textbf{0.00598} & $-16\%$ \\
137 & Far extrap & 0.00856 & \textbf{0.00601} & $-30\%$ \\
\midrule
\multicolumn{2}{@{}l}{Mean $\Delta$} & & & $-21\%$ / $-15\%$ / $-32\%$ \\
\bottomrule
\end{tabular}
\end{table}

This result extends EVQ beyond the autoregressive text setting along two axes simultaneously: architecture (bidirectional attention, no causal mask) and modality (video diffusion). The head-to-head design is particularly important because DiT training exhibits substantial CUDA non-determinism across runs, making cross-run comparisons unreliable. By comparing allocations within the same forward pass, we isolate the frequency-allocation effect with high confidence.

\paragraph{Scale-up validation: 382M DiT.}
To verify that EVQ's advantage persists at larger scale, we repeat the head-to-head experiment with a 382M DiT (${\approx}3\times$ the 129.6M model). Table~\ref{tab:dit-382m} reports the results. Training loss is identical ($+0.3\%$), confirming no capacity cost. Under YaRN extrapolation, EVQ improves far-frame MSE by $-35\%$, consistent with the 129.6M mean ($-32\%$). Without YaRN, the advantage grows to $-64\%$ (vs.\ $-59\%$ at 129.6M under raw extrapolation in Table~\ref{tab:dit-inference}), suggesting that EVQ's benefit does not diminish---and may increase---with model scale.

\begin{table}[ht]
\caption{Scale-up validation: 382M DiT head-to-head (seed 42, $4\times$ temporal extrapolation). Training loss is identical; EVQ advantage is consistent with the 129.6M results.}
\label{tab:dit-382m}
\centering
\small
\begin{tabular}{@{}l r r r@{}}
\toprule
Metric & Geo ($\tau{=}0$) & EVQ ($\tau{=}1.5$) & $\Delta$ \\
\midrule
Training loss & 0.02665 & 0.02672 & $+0.3\%$ \\
YaRN train MSE & 0.01344 & \textbf{0.00858} & $-36\%$ \\
YaRN far MSE & 0.01233 & \textbf{0.00799} & $-35\%$ \\
No-YaRN far MSE & 0.11408 & \textbf{0.04154} & $-64\%$ \\
\bottomrule
\end{tabular}
\end{table}

This result is single-seed supporting evidence, consistent with the two-seed 129.6M result. Together with the base sweep (Table~\ref{tab:dit-base-sweep}) and dead-channel validation (Table~\ref{tab:dit-base1000}), it completes a three-pronged cross-modal evidence package: EVQ transfers to bidirectional attention DiTs, is robust across base values, and scales to at least $382$M parameters.

\paragraph{Robustness across inference-time methods.}
We evaluate the same DiT checkpoints under three inference-time configurations: YaRN (progressive per-channel scaling), RIFLEx~\citep{wang2025riflex} (intrinsic-frequency reduction, official implementation with $0.9$ safety factor), and no scaling (raw extrapolation). Table~\ref{tab:dit-inference} reports far-frame denoising MSE ($t{=}0.5$). EVQ outperforms Geo under \emph{all} configurations ($-28\%$ to $-59\%$). Both inference-time methods improve over raw extrapolation, with YaRN performing best---consistent with YaRN's per-channel progressive scaling addressing more dead channels than RIFLEx's single-frequency correction. Notably, EVQ's raw performance ($0.0108$) already approaches Geo+RIFLEx ($0.0145$), suggesting that EVQ captures at training time much of the frequency correction that RIFLEx provides at inference time.

\begin{table}[ht]
\caption{DiT far-frame denoising MSE ($t{=}0.5$, $4\times$ extrapolation) under different inference-time methods. EVQ wins all configurations. YaRN is the best inference method; EVQ raw nearly matches Geo+RIFLEx.}
\label{tab:dit-inference}
\centering
\small
\begin{tabular}{@{}l r r r@{}}
\toprule
Inference method & Geo ($\tau{=}0$) & EVQ ($\tau{=}1.5$) & $\Delta$ \\
\midrule
YaRN & 0.00998 & \textbf{0.00631} & $-37\%$ \\
RIFLEx & 0.01452 & \textbf{0.01050} & $-28\%$ \\
None (raw) & 0.02625 & \textbf{0.01078} & $-59\%$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Multi-timestep validation.}
To verify that EVQ's advantage is not an artifact of the noise level chosen for evaluation, we evaluate at three diffusion timesteps (seed 42, YaRN, far-frame MSE). Table~\ref{tab:dit-timestep} shows EVQ wins at all timesteps, with the largest advantage at $t{=}0.5$ ($-36\%$). At low noise ($t{=}0.2$), where position encoding is less critical, the gap narrows to $-4\%$; at high noise ($t{=}0.8$), EVQ still wins by $-27\%$.

\begin{table}[ht]
\caption{DiT far-frame MSE across diffusion timesteps (seed 42, YaRN). EVQ wins at all noise levels.}
\label{tab:dit-timestep}
\centering
\small
\begin{tabular}{@{}l r r r@{}}
\toprule
Timestep $t$ & Geo ($\tau{=}0$) & EVQ ($\tau{=}1.5$) & $\Delta$ \\
\midrule
0.2 (low noise) & 0.00383 & \textbf{0.00366} & $-4\%$ \\
0.5 (medium) & 0.00982 & \textbf{0.00631} & $-36\%$ \\
0.8 (high noise) & 0.10368 & \textbf{0.07559} & $-27\%$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Teacher-forced temporal precision.}
To isolate position-encoding quality from autoregressive error accumulation, we evaluate the same checkpoints in teacher-forced mode on 128-frame sequences (4$\times$ extrapolation with YaRN). Table~\ref{tab:temporal-precision} reports per-region top-1 and top-5 accuracy. In the training region (frames 0--31), both models are tied to within $0.06\%$, confirming matched capacity. In the extrapolation region, EVQ leads by $+3.14\%$ top-1 and $+5.40\%$ top-5. The advantage peaks in near-extrapolation (32--63f): $+3.74\%$ top-1, $+6.59\%$ top-5.

\begin{table}[ht]
\caption{Teacher-forced accuracy on Oscillating Moving MNIST (128f, 4$\times$ YaRN, $N{=}2000$). Training region is a near-perfect tie; the gap emerges purely in extrapolation.}
\label{tab:temporal-precision}
\centering
\small
\begin{tabular}{@{}lrrrrr@{}}
\toprule
& \multicolumn{2}{c}{Top-1 Accuracy} & \multicolumn{2}{c}{Top-5 Accuracy} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Region & geo\_k16 & evq\_k16 & geo\_k16 & evq\_k16 \\
\midrule
Train (0--31f) & 74.62\% & 74.60\% & 82.07\% & 82.03\% \\
Near extrap (32--63f) & 71.99\% & \textbf{74.69\%} & 77.31\% & \textbf{82.41\%} \\
Mid extrap (64--95f) & 70.24\% & \textbf{72.89\%} & 73.54\% & \textbf{78.24\%} \\
Far extrap (96--127f) & 70.07\% & \textbf{71.39\%} & 73.10\% & \textbf{75.41\%} \\
\midrule
All extrap (32--127f) & 70.77\% & \textbf{72.99\%} & 74.65\% & \textbf{78.69\%} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Frequency-resolved decomposition.}
Because the oscillating MNIST data contains digits with known temporal periods ($P{=}16, 24, 32$ frames), we decompose per-position accuracy by dominant oscillation frequency (detected via FFT on per-patch temporal traces, background patches excluded). Table~\ref{tab:freq-decomp} shows a monotonic frequency gradient: the EVQ advantage scales with temporal frequency, from $+3.55\%$ top-1 at $P{=}32$ (lowest frequency) to $+5.07\%$ at $P{=}16$ (highest frequency). Top-5 accuracy amplifies the pattern further, reaching $+10.39\%$ in the near-extrapolation zone at $P{=}16$.

This gradient directly validates the theoretical prediction. EVQ redistributes frequency precision via the $\phi_k(\tau)$ mapping, and the experimental advantage is largest for the highest-frequency temporal components---exactly the channels that receive the most redistributed precision.

\begin{table}[ht]
\caption{Frequency-resolved top-1/top-5 accuracy delta (EVQ$-$geo) on foreground patches, YaRN mode. Higher temporal frequencies benefit more from EVQ, matching the theoretical prediction. Training region deltas are $<0.1\%$ for all periods (omitted).}
\label{tab:freq-decomp}
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
& \multicolumn{2}{c}{Top-1 $\Delta$} & \multicolumn{2}{c}{Top-5 $\Delta$} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Period & Near extrap & All extrap & Near extrap & All extrap \\
\midrule
$P{=}16$ (highest) & $+6.07\%$ & $+5.07\%$ & $\mathbf{+10.39\%}$ & $+8.48\%$ \\
$P{=}24$ (mid) & $+5.37\%$ & $+4.39\%$ & $+9.40\%$ & $+7.63\%$ \\
$P{=}32$ (lowest) & $+4.32\%$ & $+3.55\%$ & $+7.58\%$ & $+6.25\%$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Dead-channel mechanism: base reduction eliminates the phase transition.}
At base${}=10{,}000$ with $K_t{=}16$ temporal frequencies and 32 training frames, the lowest-frequency channels accumulate less than $0.1$~rad of phase rotation over the entire training sequence---they are effectively dead. EVQ's $\tau$ parameter activates these channels by redistributing frequency mass, but at base${}=10{,}000$ the activation is discontinuous: $\tau{=}1.2$ leaves 11/16 channels alive (effective rank 9), while $\tau{=}1.5$ activates 12/16 (rank 10), producing the sharp phase transition observed in Table~\ref{tab:dit-h2h}.

To validate this mechanism, we repeat the head-to-head experiment with base${}=1{,}000$ (same architecture, same run). At this base, the geometric allocation already starts with 13/16 alive channels (rank 10), so the dead-channel bottleneck is resolved. Table~\ref{tab:dit-base1000} confirms the prediction: (1)~$\tau{=}1.2$ fully recovers, winning Geo by $-48\%$ far-frame MSE (vs.\ $+162\%$ at base${}=10{,}000$); (2)~$\tau{=}1.2$ and $\tau{=}1.5$ are statistically indistinguishable ($<0.3\%$ difference on all metrics), confirming that the phase transition was caused by the dead-channel threshold, not by any intrinsic property of $\tau{=}1.2$.

\begin{table}[ht]
\caption{Dead-channel validation: base${}=1{,}000$ DiT head-to-head (129.6M, seed 42, YaRN). The sharp phase transition between $\tau{=}1.2$ and $\tau{=}1.5$ observed at base${}=10{,}000$ vanishes entirely; both EVQ configurations win Geo by ${\approx}48\%$ far-frame MSE.}
\label{tab:dit-base1000}
\centering
\small
\begin{tabular}{@{}l r r r r@{}}
\toprule
$\tau$ & Train MSE & All extrap & Far extrap & $\Delta_\text{far}$ vs Geo \\
\midrule
0.0 (Geo) & 0.02037 & 0.01274 & 0.01646 & --- \\
1.2 & \textbf{0.00933} & \textbf{0.00820} & \textbf{0.00859} & $-48\%$ \\
1.5 & \textbf{0.00930} & 0.00822 & \textbf{0.00858} & $-48\%$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Dead temporal channels across video DiT architectures.}
The dead-channel problem is not specific to our experimental setup. Table~\ref{tab:dead-channels} audits five major video DiT architectures, all of which inherit base${}=10{,}000$ from text transformers. For each model, we compute the per-channel phase rotation $\theta_k \cdot T_\text{train}$ at the training temporal length, where $\theta_k = \text{base}^{-(k+0.5)/K_t}$ is the $k$-th temporal frequency. Channels with phase $<0.1$~rad are counted as dead.

Every video model wastes 32--50\% of its temporal frequency channels, because video temporal sequences (13--51 latent tokens after VAE compression) are 80--315$\times$ shorter than text sequences (${\geq}4096$ tokens) for which base${}=10{,}000$ was designed. The text reference (LLaMA-2) has zero dead channels, confirming that the problem is specific to the sequence-length mismatch in the video temporal axis. EVQ-Cosh addresses this by redistributing frequency mass into the alive zone, and Table~\ref{tab:dit-base1000} shows that it remains effective regardless of the base choice.

\begin{table}[ht]
\caption{Dead temporal channels in major video DiT architectures. All models inherit base${}=10{,}000$ from text transformers. Channels with phase rotation $<0.1$~rad at the training temporal length are effectively dead. Video models waste 32--50\% of temporal channels; the text reference (LLaMA-2) has zero dead channels.}
\label{tab:dead-channels}
\centering
\small
\begin{tabular}{@{}l r r r r r@{}}
\toprule
Model & $K_t$ & $T_\text{train}$ & Dead / $K_t$ & Dead\% & Base oversize \\
\midrule
CogVideoX-5B & 8 & 13 & 4/8 & 50\% & $56\times$ \\
Wan-2.1 1.3B & 22 & 21 & 9/22 & 41\% & $42\times$ \\
Latte-1 & 12 & 16 & 5/12 & 42\% & $50\times$ \\
Open-Sora 1.2 & 22 & 51 & 7/22 & 32\% & $17\times$ \\
HunyuanVideo & 8 & 33 & 3/8 & 38\% & $21\times$ \\
Our DiT & 16 & 32 & 6/16 & 38\% & $26\times$ \\
\midrule
LLaMA-2 7B (text) & 64 & 4096 & 0/64 & 0\% & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Base sweep: EVQ robustness across the base spectrum.}
We sweep the temporal RoPE base across four orders of magnitude ($100$--$10{,}000$) to characterize how EVQ interacts with the dead-channel severity. Both GEO ($\tau{=}0$) and EVQ ($\tau{=}1.5$) are evaluated head-to-head within the same run at each base, under both YaRN-scaled and raw (no scaling) 4$\times$ temporal extrapolation.

\begin{table}[ht]
\caption{Base sweep: far-frame denoising MSE ($t{=}0.5$, $4\times$ extrapolation) as a function of temporal RoPE base. EVQ is remarkably stable across base${\geq}500$; GEO+YaRN exhibits non-monotonic behavior with a catastrophic spike at base${}=5{,}000$.}
\label{tab:dit-base-sweep}
\centering
\small
\begin{tabular}{@{}l r r r r r r r@{}}
\toprule
& \multicolumn{3}{c}{With YaRN} & \multicolumn{3}{c}{Without YaRN} & \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Base & GEO & EVQ & $\Delta$ & GEO & EVQ & $\Delta$ & Dead ch. \\
\midrule
100 & \textbf{0.0161} & 0.0189 & $+18\%$ & \textbf{0.296} & 0.356 & $+20\%$ & 0/16 \\
500 & 0.0210 & \textbf{0.0207} & $-1\%$ & 0.224 & \textbf{0.146} & $-35\%$ & ${\sim}3$/16 \\
1000 & 0.0517 & \textbf{0.0178} & \cellcolor{green!15}$-66\%$ & 0.286 & \textbf{0.132} & \cellcolor{green!15}$-54\%$ & ${\sim}5$/16 \\
5000 & 0.0787 & \textbf{0.0210} & \cellcolor{green!15}$-73\%$ & 0.172 & \textbf{0.111} & $-36\%$ & ${\sim}5$/16 \\
10000 & 0.0401 & \textbf{0.0137} & \cellcolor{green!15}$-66\%$ & 0.107 & \textbf{0.048} & \cellcolor{green!15}$-55\%$ & 6/16 \\
50000 & 0.1895 & \textbf{0.0262} & \cellcolor{green!15}$-86\%$ & 0.247 & \textbf{0.108} & \cellcolor{green!15}$-56\%$ & ${\sim}9$/16 \\
\bottomrule
\end{tabular}
\end{table}

Four findings emerge. \textbf{(1)}~When all channels are alive (base${}=100$, 0/16 dead), GEO wins by ${\approx}20\%$: frequency redistribution is harmful when the geometric allocation is already optimal, confirming that EVQ's advantage is specific to the dead-channel regime. \textbf{(2)}~GEO is \emph{non-monotonic} in base under both evaluation modes. With YaRN, GEO deteriorates from $0.016$ (base${}=100$) to $0.190$ (base${}=50{,}000$)---a $12\times$ range; without YaRN, GEO at base${}=1{,}000$ ($0.286$) is worse than base${}=500$ ($0.224$). This erratic behavior makes GEO's extrapolation performance unpredictable as a function of base. \textbf{(3)}~EVQ+YaRN is remarkably stable across base${\geq}500$ (MSE $\in [0.014, 0.026]$, a $1.9\times$ range vs.\ GEO's $12\times$), demonstrating intrinsic robustness to the base hyperparameter. \textbf{(4)}~The global optimum is EVQ at base${}=10{,}000$ with YaRN ($0.0137$), which is better than GEO at \emph{any} base (best GEO: $0.0161$ at base${}=100$). EVQ is a strictly superior drop-in replacement at the default base.

Without YaRN, EVQ's advantage monotonically increases with base ($-35\%$ at 500 to $-56\%$ at $50{,}000$), consistent with the dead-channel mechanism: more dead channels create more headroom for EVQ's redistribution. The fact that EVQ wins without YaRN from base${\geq}500$ confirms that its advantage is a \emph{training-time} frequency allocation effect, not an artifact of inference-time scaling.

\paragraph{Isolation experiment: pure training-length evaluation.}
To definitively isolate whether EVQ's advantage stems from training dynamics or extrapolation, we evaluate all 12 checkpoints from the base sweep on 32-frame sequences (the training length) with \emph{no} YaRN scaling---i.e., the pure training distribution.

\begin{table}[ht]
\caption{Pure training-length evaluation (32f, no YaRN, no extrapolation). All 12 configurations produce nearly identical MSE ($0.0093$--$0.0102$, $<6\%$ variation), proving that frequency allocation has negligible effect within the training distribution. The entire $+18\%$ to $-86\%$ range in Table~\ref{tab:dit-base-sweep} is a pure \emph{extrapolation} phenomenon.}
\label{tab:dit-32f-isolation}
\centering
\small
\begin{tabular}{@{}l r r r@{}}
\toprule
Base & GEO 32f MSE & EVQ 32f MSE & $\Delta$ \\
\midrule
100 & 0.00973 & 0.00961 & $-1.2\%$ \\
500 & 0.00937 & 0.00955 & $+2.0\%$ \\
1000 & 0.00940 & 0.00932 & $-0.9\%$ \\
5000 & 0.00969 & 0.00962 & $-0.7\%$ \\
10000 & 0.00987 & 0.00978 & $-0.9\%$ \\
50000 & 0.01020 & 0.00960 & $-5.9\%$ \\
\bottomrule
\end{tabular}
\end{table}

This result has three implications. First, it confirms that the model learns equally well regardless of frequency allocation: neither $\tau$ nor base affects denoising quality within the training distribution ($<6\%$ variation across all 12 configurations). Second, it proves that the catastrophic GEO failures at high base (e.g., $0.190$ at base${}=50{,}000$ with YaRN, vs.\ $0.010$ at 32f) are \emph{entirely} an extrapolation artifact---the model is well-trained but its frequency layout cannot generalize beyond the training temporal length. Third, it resolves the $\tau{=}1.2$ paradox from Table~\ref{tab:dit-h2h}: since all $\tau$ values produce identical 32f MSE, the sharp phase transition at $\tau{\in}(1.2, 1.5)$ must originate in the interaction between the frequency layout and temporal YaRN scaling, not in any deficiency of the learned model.
