\documentclass{article}

\usepackage[nonatbib]{neurips_2026}
\usepackage{natbib}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{url}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[table]{xcolor}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[draft]{hyperref}
\else
  \usepackage{hyperref}
\fi
\usepackage{placeins}

\renewcommand{\ttdefault}{pcr}

\graphicspath{{figs/}}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  breaklinks=true,
}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

\title{EVQ-Cosh: Variational Frequency Allocation for Rotary Position Embedding}

\author{Anonymous Authors}

\begin{document}

\maketitle

\begin{abstract}
The default RoPE frequency table is not the RoPE operator; it is an inherited allocation of a scarce spectral budget. We study training-time frequency allocation as a RoPE design axis complementary to operator design and inference-time scaling. From a stated broadband collision surrogate over log-frequency densities, we derive \textsc{EVQ-Cosh}: a closed-form inverse-CDF allocation for the same finite set of RoPE channels. It changes only the inverse-frequency initialization, adds zero learned parameters, and uses $\tau=d_{\mathrm{head}}/\sqrt{L}$ (or architecture-specific $d_{\mathrm{eff}}/\sqrt{L}$) as an operating default inside an empirical basin rather than as a claimed global optimum. Mechanism stress tests show that allocation shape changes how range scaling and compressed-RoPE architectures use their channels: under matched YaRN scale on a $454$M passkey-mix model, EVQ+YaRN reaches $100\%$ passkey retrieval at $8\mathrm{K}$ while Geo+YaRN remains at $61\%$ across three seeds; in a $432$M MLA model with only $16$ rotary channels, EVQ alone reduces $2\times$ extrapolation PPL by $31.1\%$ at $+1.1\%$ in-distribution cost. PE-dominant and supporting tests further indicate that the geometric schedule is not a neutral implementation detail, but a substantive inductive bias for long-context behavior.
\end{abstract}

\input{sections/01_intro}
\input{sections/02_related}
\input{sections/03_theory}
\input{sections/05_experiments}
\input{sections/06_limitations}

\newpage
\bibliographystyle{plainnat}
\bibliography{refs/references}

\clearpage
\appendix

\input{appendix/a1_proofs}
\input{appendix/a2_experiment_details}
\input{appendix/a3_supporting_results}
\input{appendix/a4_supporting_experiments}

\section*{Broader Impact Statement}
\label{sec:broader-impact}

This work studies frequency allocation as a design axis for rotary position embedding rather than a specific deployed application. Its main positive impact is scientific and systems-oriented: replacing the default geometric frequency schedule with a principled closed-form allocation can reduce wasted compute and improve the reliability of long-context models in tested RoPE-based settings. The main risk is over-generalization: our strongest evidence is concentrated on text extrapolation and retrieval-heavy long-context evaluation, so we do not present EVQ as a universal guarantee of downstream improvement across all tasks, modalities, or deployment settings.

\clearpage
\section*{NeurIPS Paper Checklist}

\begin{enumerate}
\item \textbf{Claims}
\begin{itemize}
\item \textbf{Question:} Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The abstract and introduction are restricted to three contributions: (1) identifying training-time frequency allocation as a PE design axis with a closed-form variational family, (2) demonstrating that a semi-analytic rule---with derived $d_{\mathrm{head}}$ and $L^{-1/2}$ structure and an empirically calibrated $\mathcal{O}(1)$ scale---falls within the best observed sweep basin across the tested grid of 27 configurations (multi-seed anchors support the systems and MLA claims; the PE-dominant DAPE-style diagnostic is explicitly seed-42 for Geo/DAPE/EVQ), and (3) the matched-scale systems result that EVQ increases YaRN leverage under a fixed YaRN scale (we do not claim dominance over every tuned YaRN baseline). The $\tau^*$ rule's epistemic status---derived $d_{\mathrm{head}}$ and $L^{-1/2}$ structure under the stated small-$\tau$ scalings, channel-load-motivated $\chi^2$ stiffness, and unit-convention $\lambda$---is stated explicitly in \S3.7.
\end{itemize}

\item \textbf{Limitations}
\begin{itemize}
\item \textbf{Question:} Does the paper discuss the limitations of the work performed by the authors?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The body explicitly marks the broadband surrogate as the main approximation, presents \(\tau^*\) as a semi-analytic operating-point rule rather than a tuned hyperparameter, and demotes single-seed and video evidence to supporting status.
\end{itemize}

\item \textbf{Theory assumptions and proofs}
\begin{itemize}
\item \textbf{Question:} For each theoretical result, does the paper provide the full set of assumptions and a complete (and correct) proof?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} Theorems~1--2 and Proposition~1 are exact conditional on the broadband surrogate; the surrogate itself is supported by a functional check (24--92\% collision-score reduction across 12 configurations) in Appendix~\ref{sec:surrogate-validation}. The $\tau^*$ scaling law (Proposition~\ref{prop:softmax-transport}) is semi-analytic: $d_{\mathrm{head}}$ dependence and the $L^{-1/2}$ exponent are derived, but one transport constant $\lambda$ is calibrated empirically. This status is stated explicitly in \S3.7 and the limitations section.
\end{itemize}

\item \textbf{Experimental result reproducibility}
\begin{itemize}
\item \textbf{Question:} Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The body states the principal settings and the appendix records datasets, seeds, evaluation grouping, and the status of larger supporting runs.
\end{itemize}

\item \textbf{Open access to data and code}
\begin{itemize}
\item \textbf{Question:} Does the paper provide open access to the data and code?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} We include an anonymous supplementary code archive containing the EVQ-Cosh inverse-frequency initializer, the collision-surrogate and $\tau^*$-rule analyses, evaluation scripts, and configuration files for the primary experiments (EVQ$\times$YaRN 454M, PE-dominant 125M/454M, MLA 432M). Upstream datasets (FineWeb-Edu, TinyStories, QuALITY, RULER passkey/NIAH) are available from their respective public releases; our passkey-mix composition and data-prep scripts are in the archive. Model checkpoints and weights are not included to preserve double-blind review and to keep the archive within size limits. A reproducibility README lists the commands to regenerate each primary-tier table and figure from a fresh checkout.
\end{itemize}

\item \textbf{Experimental setting/details}
\begin{itemize}
\item \textbf{Question:} Does the paper specify all the training and test details?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} Main settings are given in the body and additional details are placed in the appendix.
\end{itemize}

\item \textbf{Experiment statistical significance}
\begin{itemize}
\item \textbf{Question:} Does the paper report error bars or equivalent multi-run evidence?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The matched-scale EVQ$\times$YaRN and MLA claims are anchored in 3-seed evidence. The PE-dominant DAPE-style result is intentionally reported as a seed-42 mechanism diagnostic for Geo/DAPE/EVQ, while single- and two-seed scope extensions are labeled as supporting or exploratory.
\end{itemize}

\item \textbf{Experiments compute resources}
\begin{itemize}
\item \textbf{Question:} For each experiment, does the paper provide sufficient information on the computer resources used?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The appendix records model scales, training token counts, context windows, batch sizes, and seed counts for every primary run. Primary experiments were conducted on an anonymous internal A100 / H100 accelerator cluster; total compute for the primary tier (454M EVQ$\times$YaRN + 432M MLA + 125M/454M PE-dominant) is on the order of a few hundred accelerator-hours, with the 454M/432M anchors dominating. Supporting LoRA-on-8B and video-DiT rows each use at most tens of accelerator-hours. Exact wall-clock varies with cluster scheduling and is not used as evidence for the scientific comparisons.
\end{itemize}

\item \textbf{Code of ethics}
\begin{itemize}
\item \textbf{Question:} Does the research conducted in the paper conform to the NeurIPS Code of Ethics?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The work studies frequency allocation for rotary position embedding and does not involve human-subject experimentation or a deployed user-facing system.
\end{itemize}

\item \textbf{Broader impacts}
\begin{itemize}
\item \textbf{Question:} Does the paper discuss both potential positive societal impacts and negative societal impacts of the work performed?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} The limitations discussion and checklist justifications state both the potential systems benefits of better long-context positional design and the risk of over-generalizing beyond the present evidence.
\end{itemize}

\item \textbf{Safeguards}
\begin{itemize}
\item \textbf{Question:} Does the paper describe safeguards that have been put in place for responsible release?
\item \textbf{Answer:} \answerNA{}
\item \textbf{Justification:} The submission does not release a deployed system or a safety-sensitive product.
\end{itemize}

\item \textbf{Licenses}
\begin{itemize}
\item \textbf{Question:} Are the creators or original owners of assets used in the paper properly credited?
\item \textbf{Answer:} \answerYes{}
\item \textbf{Justification:} External methods, benchmarks, and model families are cited in the references.
\end{itemize}

\item \textbf{New assets}
\begin{itemize}
\item \textbf{Question:} Are new assets introduced in the paper well documented?
\item \textbf{Answer:} \answerNA{}
\item \textbf{Justification:} The submission does not introduce a released dataset or benchmark asset.
\end{itemize}

\item \textbf{Crowdsourcing and human subjects}
\begin{itemize}
\item \textbf{Question:} For crowdsourcing experiments and research with human subjects, does the paper include the full text of instructions given to participants?
\item \textbf{Answer:} \answerNA{}
\item \textbf{Justification:} No crowdsourcing or human-subject research is involved.
\end{itemize}

\item \textbf{IRB approvals}
\begin{itemize}
\item \textbf{Question:} Does the paper describe potential risks incurred by study participants?
\item \textbf{Answer:} \answerNA{}
\item \textbf{Justification:} No human-subject study is involved.
\end{itemize}

\item \textbf{Declaration of LLM usage}
\begin{itemize}
\item \textbf{Question:} Does the paper describe the usage of LLMs if it is an important, original, or non-standard component of the core methods in this research?
\item \textbf{Answer:} \answerNA{}
\item \textbf{Justification:} LLMs and agents were used only for author-side writing, editing, code assistance, and submission-hygiene checks. They are not an important, original, or non-standard component of the EVQ-Cosh method, theoretical derivation, experiments, or evaluation protocol.
\end{itemize}
\end{enumerate}

\end{document}
